Simulating High-Dimensional Multivariate Data using the \pkg{Bigsimr} Package

### 2021-04-01 article_bigsimr.tex

\author{
\\
}
\title{Simulating High-Dimensional Multivariate Data using the \pkg{Bigsimr} Package}

\Plainauthor{}
\Plaintitle{Simulating High-Dimensional Multivariate Data using the Bigsimr Package}
\Shorttitle{\pkg{Bigsimr}: Simulate High-Dimensional Data}

@@

\Address{
  }

@@

% Pandoc header
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{ragged2e}

\usepackage{setspace}
\setstretch{2.0}

@@

We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For ease of modeling and simplicity of exposition, we only consider high expressing genes. In turn, we begin by filtering to retain the top 1000 of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from \(N=878\) patients' tumor samples. This gives a great number of pairwise dependencies among the marginals (specifically, \(\ensuremath{4.995\times 10^{5}}\) correlation parameters). Table \ref{tab:ch010-realDataTab} displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples. To help visualize the bivariate relationships for these three selected genes across all patients, Figure \ref{fig:ch010-realDataFig} displays the marginal distributions and estimated Spearman's correlations.

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2020-02-25)
R> num_genes    <- 3
R> num_patients <- 5
R> gene_sample  <- sample(example_genes[1:1000], num_genes)
R> 
R> cap <- paste0("mRNA expression for three selected high-expressing genes, ",
+               paste(gene_sample, collapse = ", "), 
+               ", for the first five patients in the TCGA BRCA data set.")
R> 
R> small_brca <- example_brca %>%
+   select(all_of(gene_sample)) %>%
+   head(n = num_patients) %>%
+   as_tibble(rownames = "Patient ID") %>%
+   mutate(`Patient ID` = str_sub(`Patient ID`, end = 12))
R> 
R> small_brca %>%
+   kable(format = "latex", booktabs = TRUE, caption = cap)
\end{CodeInput}
\begin{table}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> ggpairs(
+   data = example_brca[, gene_sample],
+   upper = list(continuous = wrap('cor', method = "spearman"))
+ ) + 
+   theme_bw() +
+   theme(axis.text.x = element_text(angle = -90, vjust = 0.5))
\end{CodeInput}
\begin{CodeOutput}
Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
exact p-value with ties
\end{CodeOutput}
\begin{CodeOutput}
Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
exact p-value with ties
\end{CodeOutput}
\begin{CodeOutput}
Warning in cor.test.default(x, y, method = method, use = use): Cannot compute
exact p-value with ties
\end{CodeOutput}
\begin{figure}

{\centering \includegraphics{article_bigsimr_files/figure-latex/ch010-realDataFig-1}

@@

\begin{equation}
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
\label{eq:pearson}
\end{equation}

@@

\begin{equation}
\label{eq:convertKendall}
\rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

@@

\begin{equation}
\label{eq:convertSpearman}
\rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

@@

\begin{equation}
\label{eq:frechet}
\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
\end{equation}

\noindent where \(U\) is a uniform random variable on \((0,1)\) and \(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\), respectively, definitely by \eqref{eq:inverseCDF} when the variables are discrete.

\begin{equation}
F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
\label{eq:inverseCDF}
\end{equation}

@@

\begin{equation}
F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
\label{eq:copula}
\end{equation}

@@

\begin{equation}
\label{eq:gauss}
F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

@@

This section describes our methods involved in simulating a random vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\). Each \(Y_i\) has a specified marginal CDF \(F_i\) and its inverse \(F^{-1}_i\). To characterize dependency, every pair \((Y_i, Y_j)\) has a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\), and/or Kendall's \(\tau\). The method can be described as a \emph{high-performance Gaussian copula} (Equation \eqref{eq:gauss}) providing a high-dimensional NORTA-inspired algorithm.

@@

\begin{equation}
    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
    \label{eq:pearsonIntegralRelation}
\end{equation}

@@

The first two descriptions of the \emph{Mapping step} employ the closed-form relationships between \(\rho_S\) and \(\tau\) with \(\rho_P\) for bivariate normal random variables via Equations \eqref{eq:convertKendall} and \eqref{eq:convertSpearman}, respectively (implemented as \texttt{cor\_covert}). Initializing our algorithm to match the nonparametric correlations by computing these equations for all pairs is computationally trivial.

For the computationally expensive process to match Pearson correlations, we approximate Equation \eqref{eq:pearsonIntegralRelation} for all pairs of margins. To this end, we implement the approximation scheme introduced by \citet{xiao2019matching}. Briefly, the many double integrals of the form in \eqref{eq:pearsonIntegralRelation} are approximated by weighted sums of Hermite polynomials. Matching coefficients for pairs of continuous distributions is made tractable by this method, but for discrete distributions (especially discrete distributions with large support sets or infinite support), the problem is not so simple or efficient.

@@

There is an increasing probability of encountering a non-admissible correlation matrix as dimension increases. In our experience, the mapping step for large \(d\) almost always produces a \(R_X\) that is not a correlation matrix. In Section \ref{package} we provide a basic description of how to quantify and control the approximation error. Further, the RNA-seq example in Section \ref{examples} provides an illustration of this in practice.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> box::use(
+   patchwork[...],
+   dplyr[...],
+   ggplot2[...],
+   tidyr[drop_na],
+   bigsimr[bigsimr_setup, distributions_setup],
+   JuliaCall[julia_call]
+ )
R> 
R> Sys.setenv(JULIA_NUM_THREADS = parallel::detectCores())
R> bs <- bigsimr_setup(pkg_check = FALSE)
\end{CodeInput}
\begin{CodeOutput}
Julia version 1.5.4 at location /home/alex/julia-1.5.4/bin will be used.
\end{CodeOutput}
\begin{CodeOutput}
Loading setup script for JuliaCall...
\end{CodeOutput}
\begin{CodeOutput}
Finish loading setup script for JuliaCall.
\end{CodeOutput}
\begin{CodeInput}
R> dist <- distributions_setup()
\end{CodeInput}
\end{CodeChunk}

This section describes a low-dimensional (2D) random vector simulation workflow via the \texttt{bigsimr} \texttt{R} package (\url{https://github.com/SchisslerGroup/r-bigsimr}). The package \texttt{bigsimr} provides an interface to the native code written in Julia, registered as the \texttt{Bigsimr} Julia package (\url{https://github.com/adknudson/Bigsimr.jl}). In addition to the native Julia \texttt{Bigsimr} package and \texttt{R} interface \texttt{bigsimr}, we also provide a python interface \texttt{bigsimr} (\url{https://github.com/SchisslerGroup/python-bigsimr/}) that interfaces with the Julia \texttt{Bigsimr} package. The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \ref{algorithms}).

@@

\begin{CodeChunk}
\begin{CodeInput}
R> glimpse(df)
\end{CodeInput}
\begin{CodeOutput}
Rows: 116
Columns: 2
$ Temp  <int> 67, 72, 74, 62, 66, 65, 59, 61, 74, 69, 66, 68, 58, 64, 66, 57, ~
$ Ozone <int> 41, 36, 12, 18, 28, 23, 19, 8, 7, 16, 11, 14, 18, 14, 34, 6, 30,~
\end{CodeOutput}
\end{CodeChunk}

Figure \ref{fig:ch030-aq-joint-dist} visualizes the bivariate relationship between Ozone and Temperature. We aim to simulate random two-component vectors mimicking this structure. The margins are not normally distributed; particularly the Ozone level exhibits a strong positive skew.

\begin{CodeChunk}
\begin{CodeInput}
R> p0 <- ggplot(df, aes(Temp, Ozone)) +
+   geom_point(size = 1) +
+   theme(legend.position = "none") + 
+   labs(x = "Temperature")
R> 
R> pTemp <- ggplot(df, aes(Temp)) + 
+   geom_density() +
+   theme(axis.title.x = element_blank(),
+         axis.title.y = element_blank(),
+         axis.text = element_blank(),
+         axis.ticks = element_blank())
R> 
R> pOzone <- ggplot(df, aes(Ozone)) + 
+     geom_density() +
+   theme(axis.title.y = element_blank(),
+         axis.title.x = element_blank(),
+         axis.ticks = element_blank(),
+         axis.text = element_blank()) +
+   coord_flip()
R> 
R> pTemp + plot_spacer() + p0 + pOzone + 
+   plot_layout(widths = c(3,1), heights = c(1, 3)) 
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{article_bigsimr_files/figure-latex/ch030-aq-joint-dist-1}

@@

Based on the estimated densities in Figure \ref{fig:ch030-aq-joint-dist}, we assume \texttt{Temp} is normally distributed and \texttt{Ozone} is log-normally distributed, as the latter values are positive and skewed. We use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:

@@

First we use \texttt{cor\_bounds} to ensure that the pairwise target correlation values are valid prior to the mapping step which constructs \(R_X\) for the MVN input into NORTA (see Section \ref{algorithms}). \texttt{cor\_bounds} estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of \citet{DH2011}.

\begin{CodeChunk}
\begin{CodeInput}
R> bounds <- bs$cor_bounds(margins)
R> bounds$lower
\end{CodeInput}
\begin{CodeOutput}
        [,1]    [,2]
[1,]  1.0000 -0.8902
[2,] -0.8902  1.0000
\end{CodeOutput}
\begin{CodeInput}
R> bounds$upper
\end{CodeInput}
\begin{CodeOutput}
       [,1]   [,2]
[1,] 1.0000 0.8876
[2,] 0.8876 1.0000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> bs$cor_bounds(margins[1], margins[2], bs$Pearson, n_samples = 1e6)
\end{CodeInput}
\begin{CodeOutput}
Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
(lower = -0.8822730883942045, upper = 0.8833146839651883)
\end{CodeOutput}
\end{CodeChunk}

@@

Figure \ref{fig:ch030-plot-sim} plots the 10,000 simulated points.

\begin{CodeChunk}
\begin{CodeInput}
R> p1 <- df_sim %>%
+   ggplot(aes(Temp, Ozone)) +
+   geom_density_2d_filled() +
+   theme(legend.position = "none") +
+   labs(x = "Simulated Temperature", y = "Simulated Ozone") +
+   scale_y_continuous(limits = c(0, max(df$Ozone))) +
+   scale_x_continuous(limits = range(df$Temp))
R> 
R> p1Temp <- ggplot(df_sim, aes(Temp)) + 
+   geom_density(alpha = 0.5, fill = "lightseagreen") +
+   theme(axis.title.x = element_blank(),
+         axis.title.y = element_blank(),
+         axis.text = element_blank(),
+         axis.ticks = element_blank())
R> 
R> p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
+   geom_density(alpha = 0.5, fill = "lightseagreen") + 
+   theme(axis.title.y = element_blank(),
+         axis.title.x = element_blank(),
+         axis.ticks = element_blank(),
+         axis.text = element_blank()) +
+   coord_flip() 
R> 
R> p1Temp + plot_spacer() + p1 + p1Ozone + 
+   plot_layout(widths = c(3,1), heights = c(1, 3))
\end{CodeInput}
\begin{CodeOutput}
Warning: Removed 342 rows containing non-finite values (stat_density2d_filled).
\end{CodeOutput}
\begin{figure}

{\centering \includegraphics{article_bigsimr_files/figure-latex/ch030-plot-sim-1}

@@

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example, and, so we proceed in increasing complexity, leading to the model in our motivating application in Section \ref{examples}. We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson's, Spearman's, and Kendall's --- in identical, bivariate marginal configurations. For each pair of identical margins, we vary the target correlation across \(\Omega\), the set of possible admissible values for correlation type, to evaluate the simulation's ability to obtain the theoretic bounds. The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (mimicking RNA-seq counts).

Table \ref{tab:sims} lists our identical-marginal, bivariate simulation configurations. We increase the simulate replicates \(B\) to check that our results converge to the target correlations and gauge statistical efficiency. We select distributions beginning with a standard multivariate normal (MVN) as we expect the performance to be exact (up to MC error) for all correlation types. Then, we select a non-symmetric continuous distribution: a standard (rate =1) two-component multivariate gamma (MVG). Finally, we select distributions and marginal parameter values that are motivated by our RNA-seq data, namely values proximal to probabilities and sizes estimated from the data (see \href{examples}{Example applications} for estimation details). Thus we arrive at a multivariate negative binomial (MVNB) \(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).

Table: \label{tab:sims} Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> bivariate_normal_sims %>%
+     ggplot(aes(rho, rho_hat, color = type)) +
+     geom_point() +
+     geom_abline(slope = 1) +
+     facet_wrap(~ type + N) + 
+     theme_bw()
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> bivariate_gamma_sims %>%
+     ggplot(aes(rho, rho_hat, color = type)) +
+     geom_point() +
+     geom_abline(slope = 1) +
+     facet_wrap(~ type + N) + 
+     theme_bw()
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> bivariate_nbinom_sims %>%
+     ggplot(aes(rho, rho_hat, color = type)) +
+     geom_point() +
+     geom_abline(slope = 1) +
+     facet_wrap(~ type + N) + 
+     theme_bw()
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> allDat <- bind_rows(
+     select(bivariate_normal_sims, margins, type, N, rho, rho_hat),
+     select(bivariate_gamma_sims, margins, type, N, rho, rho_hat),
+     select(bivariate_nbinom_sims, margins, type, N, rho, rho_hat)
+ ) %>%
+     mutate(margins = factor(margins, levels=c("norm", "gamma", "nbinom")),
+            type = factor(type, levels=c("Pearson", "Spearman", "Kendall")),
+            N = factor(N))
\end{CodeInput}
\end{CodeChunk}

Figure \ref{fig:ch040-bPlot} displays the aggregated bivariate simulation results. Table \ref{tab:ch040-BiError} contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.

\begin{CodeChunk}
\begin{CodeInput}
R> tabMAE <- allDat %>%
+     group_by(N, type, margins) %>%
+     summarize(MAE = mean(abs(rho - rho_hat))) %>%
+     ungroup()
\end{CodeInput}
\begin{CodeOutput}
`summarise()` has grouped output by 'N', 'type'. You can override using the `.groups` argument.
\end{CodeOutput}
\begin{CodeInput}
R> kable(tabMAE, booktabs = TRUE, format = "latex",
+       linesep = c("", "", "\\addlinespace"),
+       col.names = c("No. of random vectors",
+                     "Correlation type",
+                     "Distribution",
+                     "Mean abs. error"),
+       caption = "Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.")
\end{CodeInput}
\begin{table}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
R> # New facet label names
R> repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
R> names(repsLabs) <- c(1000, 10000, 100000)
R> typeLabs <- c("Pearson", "Spearman", "Kendall")
R> names(typeLabs) <- c("Pearson", "Spearman", "Kendall")
R> 
R> # Set colors
R> ## RColorBrewer::display.brewer.all()
R> numColors <- 4
R> numGroups <- length(levels(allDat$margins))
R> myColors <- rev(RColorBrewer::brewer.pal(n = numColors, name = "Greys")[ ((numColors - numGroups) + 1):numColors])
R> 
R> allDat %>%
+     ggplot(aes(x = rho, y = rho_hat, color = margins)) +
+     geom_point(size = 2) +
+     scale_color_manual(values = myColors ) +
+     geom_abline(slope = 1, linetype = "dashed") +
+     labs(x = "Specified Correlation", y = "Estimated Correlation") +
+     facet_wrap(~ type + N, labeller = labeller(N = repsLabs, type = typeLabs)) +
+     theme_bw() +
+     theme(legend.position = "bottom", legend.direction = "horizontal")
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{article_bigsimr_files/figure-latex/ch040-bPlot-1}

@@

The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, with all 16 threads employed during computation. The results for \(d \leq 500\) are fast with all times (except Kendall) under 3 seconds. Table \ref{tab:ch040-moderateDtab} displays the total computation time of the algorithms steps 1 through 3, including estimation step i.

\begin{CodeChunk}
\begin{CodeInput}
R> dat_long <- benchmark_dependences %>%
+   select(-total_time, -n_sim, -needed_near_pd) %>%
+   pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
+   mutate(dim = factor(dim),
+          Step = factor(Step,
+                        levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
+                        labels = c("Compute Correlation",
+                                   "Adjust Correlation",
+                                   "Check Admissibility",
+                                   "Simulate Data"))) %>%
+     rename(Correlation = corr_type, Dimensions = dim)
R> 
R> tabTimes  <- dat_long %>%
+     filter(Dimensions %in% c(100, 250, 500)) %>%
+     group_by(Dimensions, Correlation) %>%
+     summarize('Total Time(Seconds)' = sum(Time)) %>%
+     ungroup()
\end{CodeInput}
\begin{CodeOutput}
`summarise()` has grouped output by 'Dimensions'. You can override using the `.groups` argument.
\end{CodeOutput}
\begin{CodeInput}
R> tabTimes %>%
+     kable(booktabs = TRUE,
+           linesep = c("", "", "\\addlinespace"),
+           col.names = c("Dimension",
+                         "Correlation type",
+                         "Total Time (Seconds)"),
+           caption = 'Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.')
\end{CodeInput}
\begin{table}

\caption{\label{tab:ch040-moderateDtab}Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
\centering
\begin{tabular}[t]{llr}
\toprule
Dimension & Correlation type & Total Time (Seconds)\\
\midrule
100 & Pearson & 0.080\\
100 & Spearman & 0.036\\
100 & Kendall & 0.116\\
\addlinespace
250 & Pearson & 0.286\\
250 & Spearman & 0.371\\
250 & Kendall & 0.366\\
\addlinespace
500 & Pearson & 2.932\\
500 & Spearman & 0.176\\
500 & Kendall & 2.183\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> d <- 1000
R> brca1000 <- example_brca %>%
+   select(all_of(1:d)) %>%
+   mutate(across(everything(), as.double))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> include_graphics('fig/ch050-figBRCA.png')
\end{CodeInput}
\begin{figure}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> frobenius_loss <- sapply(simRho, function(R) sum((R - R_S)^2))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> knitr::include_graphics('images/hex-bigsimr.png')
\end{CodeInput}

@@

As an invited speaker at CMStatistics 2020, I was delighted to learn of the opportunity to have our work included in the first issue of CSDA Annals of Statistical Data Science.

>>>>

\author{
Alfred G. Schissler\\University of Nevada \And Edward J.
Bedrick\\University of Nevada \And Alexander D. Knudson\\University of
Nevada \AND Tomasz J. Kozubowski\\University of Nevada \And Tin
Nguyen\\University of Nevada \And Anna K. Panorska\\University of Nevada
\AND Juli Petereit\\University of Nevada \And Walter W.
Piegorsch\\University of Nevada \And Duc Tran\\University of Nevada
}
\title{Simulating High-Dimensional Multivariate Data using the
\pkg{Bigsimr} Package}

\Plainauthor{Alfred G. Schissler, Edward J. Bedrick, Alexander D.
Knudson, Tomasz J. Kozubowski, Tin Nguyen, Anna K. Panorska, Juli
Petereit, Walter W. Piegorsch, Duc Tran}
\Plaintitle{Simulating High-Dimensional Multivariate Data using the
Bigsimr Package}
\Shorttitle{\pkg{Bigsimr}: Simulate High-Dimensional Data}

@@

\Address{
    Alfred G. Schissler\\
    University of Nevada\\
    1664 N Virginia St.\\
Reno, NV 89557\\
  E-mail: \email{aschissler@unr.edu}\\

@@

% Pandoc header

\usepackage{amsmath} \usepackage{amssymb} \usepackage{amsfonts} \usepackage{amsthm} \newtheorem{theorem}{Theorem} \newtheorem{lemma}[theorem]{Lemma} \usepackage{booktabs} \usepackage{longtable} \usepackage{array} \usepackage{ragged2e} \usepackage{setspace}

@@

We illustrate our methodology using a well-studied Breast Invasive
Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see
Acknowledgments). For ease of modeling and simplicity of exposition, we
only consider high expressing genes. In turn, we begin by filtering to
retain the top 1000 of the highest-expressing genes (in terms of median
expression) of the over 20,000 gene measurements from \(N=878\)
patients' tumor samples. This gives a great number of pairwise
dependencies among the marginals (specifically,
\(\ensuremath{4.995\times 10^{5}}\) correlation parameters). Table
@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected
high-expressing genes for the first five patients' breast tumor samples.
To help visualize the bivariate relationships for these three selected
genes across all patients, Figure @ref(fig:ch010-realDataFig) displays
the marginal distributions and estimated Spearman's correlations.

\begin{CodeChunk}
\begin{table}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch010-realDataFig-1}

@@

\begin{equation}
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
(\#eq:pearson)
\end{equation}

@@

\begin{equation}
(\#eq:convertKendall)
\rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

@@

\begin{equation}
(\#eq:convertSpearman)
\rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

@@

\begin{equation}
(\#eq:frechet)
\rho_P^{max} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho_P^{min} = \rho_P \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
\end{equation}

\noindent where \(U\) is a uniform random variable on \((0,1)\) and
\(F^{-1}_1, F^{-1}_2\) are the inverse CDFs of \(X_1\) and \(X_2\),
respectively, definitely by @ref(eq:inverseCDF) when the variables are
discrete.

\begin{equation}
F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
(\#eq:inverseCDF)
\end{equation}

@@

\begin{equation}
F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
(\#eq:copula)
\end{equation}

@@

\begin{equation}
(\#eq:gauss)
F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

@@

This section describes our methods involved in simulating a random
vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\). Each
\(Y_i\) has a specified marginal CDF \(F_i\) and its inverse
\(F^{-1}_i\). To characterize dependency, every pair \((Y_i, Y_j)\) has
a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\),
and/or Kendall's \(\tau\). The method can be described as a
\emph{high-performance Gaussian copula} (Equation @ref(eq:gauss))
providing a high-dimensional NORTA-inspired algorithm.

@@

\begin{equation}
    \mathrm{E}\left[Y_i Y_j\right] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} F_i^{-1}\left[\Phi(z_i)\right] F_j^{-1}\left[\Phi(z_j)\right] \phi(z_i, z_j, \rho_z) dz_i dz_j,
    (\#eq:pearsonIntegralRelation)
\end{equation}

@@

The first two descriptions of the \emph{Mapping step} employ the
closed-form relationships between \(\rho_S\) and \(\tau\) with
\(\rho_P\) for bivariate normal random variables via Equations
@ref(eq:convertKendall) and @ref(eq:convertSpearman), respectively
(implemented as \texttt{cor\_covert}). Initializing our algorithm to
match the nonparametric correlations by computing these equations for
all pairs is computationally trivial.

For the computationally expensive process to match Pearson correlations,
we approximate Equation @ref(eq:pearsonIntegralRelation) for all pairs
of margins. To this end, we implement the approximation scheme
introduced by \citet{xiao2019matching}. Briefly, the many double
integrals of the form in @ref(eq:pearsonIntegralRelation) are
approximated by weighted sums of Hermite polynomials. Matching
coefficients for pairs of continuous distributions is made tractable by
this method, but for discrete distributions (especially discrete
distributions with large support sets or infinite support), the problem
is not so simple or efficient.

@@

There is an increasing probability of encountering a non-admissible
correlation matrix as dimension increases. In our experience, the
mapping step for large \(d\) almost always produces a \(R_X\) that is
not a correlation matrix. In Section @ref(package) we provide a basic
description of how to quantify and control the approximation error.
Further, the RNA-seq example in Section @ref(examples) provides an
illustration of this in practice.

@@

This section describes a low-dimensional (2D) random vector simulation
workflow via the \texttt{bigsimr} \texttt{R} package
(\url{https://github.com/SchisslerGroup/r-bigsimr}). The package
\texttt{bigsimr} provides an interface to the native code written in
Julia, registered as the \texttt{Bigsimr} Julia package
(\url{https://github.com/adknudson/Bigsimr.jl}). In addition to the
native Julia \texttt{Bigsimr} package and \texttt{R} interface
\texttt{bigsimr}, we also provide a python interface \texttt{bigsimr}
(\url{https://github.com/SchisslerGroup/python-bigsimr/}) that
interfaces with the Julia \texttt{Bigsimr} package. The Julia package
provides a high-performance implementation of our proposed random vector
generation algorithm and associated functions (see Section
@ref(algorithms)).

@@

\begin{CodeChunk}
\begin{CodeOutput}
Rows: 116
Columns: 2
$ Temp  <int> 67, 72, 74, 62, 66, 65, 59, 61, 74, 69, 66, 68, 58, 64, 66, 57, ~
$ Ozone <int> 41, 36, 12, 18, 28, 23, 19, 8, 7, 16, 11, 14, 18, 14, 34, 6, 30,~
\end{CodeOutput}
\end{CodeChunk}

Figure @ref(fig:ch030-aq-joint-dist) visualizes the bivariate
relationship between Ozone and Temperature. We aim to simulate random
two-component vectors mimicking this structure. The margins are not
normally distributed; particularly the Ozone level exhibits a strong
positive skew.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch030-aq-joint-dist-1}

@@

Based on the estimated densities in Figure
@ref(fig:ch030-aq-joint-dist), we assume \texttt{Temp} is normally
distributed and \texttt{Ozone} is log-normally distributed, as the
latter values are positive and skewed. We use the well-known, unbiased
estimators for the normal distribution's parameters and maximum
likelihood estimators for the log-normal parameters:

@@

First we use \texttt{cor\_bounds} to ensure that the pairwise target
correlation values are valid prior to the mapping step which constructs
\(R_X\) for the MVN input into NORTA (see Section @ref(algorithms)).
\texttt{cor\_bounds} estimates the pairwise lower and upper correlation
bounds using the Generate, Sort, and Correlate algorithm of
\citet{DH2011}.

\begin{CodeChunk}
\begin{CodeInput}
R> bounds <- bs$cor_bounds(margins)
R> bounds$lower
\end{CodeInput}
\begin{CodeOutput}
        [,1]    [,2]
[1,]  1.0000 -0.8745
[2,] -0.8745  1.0000
\end{CodeOutput}
\begin{CodeInput}
R> bounds$upper
\end{CodeInput}
\begin{CodeOutput}
       [,1]   [,2]
[1,] 1.0000 0.8748
[2,] 0.8748 1.0000
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> bs$cor_bounds(margins[1], margins[2], bs$Pearson, n_samples = 1e6)
\end{CodeInput}
\begin{CodeOutput}
Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
(lower = -0.8799819883897544, upper = 0.8798667127065275)
\end{CodeOutput}
\end{CodeChunk}

@@

Figure @ref(fig:ch030-plot-sim) plots the 10,000 simulated points.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch030-plot-sim-1}

@@

We select bivariate simulation configurations to ultimately simulate our
motivating discrete-valued RNA-seq example, and, so we proceed in
increasing complexity, leading to the model in our motivating
application in Section @ref(examples). We begin with empirically
evaluating the dependency matching across all three supported
correlations --- Pearson's, Spearman's, and Kendall's --- in identical,
bivariate marginal configurations. For each pair of identical margins,
we vary the target correlation across \(\Omega\), the set of possible
admissible values for correlation type, to evaluate the simulation's
ability to obtain the theoretic bounds. The simulations progress from
bivariate normal, to bivariate gamma (non-normal yet continuous), and
bivariate negative binomial (mimicking RNA-seq counts).

Table @ref(tab:sims) lists our identical-marginal, bivariate simulation
configurations. We increase the simulate replicates \(B\) to check that
our results converge to the target correlations and gauge statistical
efficiency. We select distributions beginning with a standard
multivariate normal (MVN) as we expect the performance to be exact (up
to MC error) for all correlation types. Then, we select a non-symmetric
continuous distribution: a standard (rate =1) two-component multivariate
gamma (MVG). Finally, we select distributions and marginal parameter
values that are motivated by our RNA-seq data, namely values proximal to
probabilities and sizes estimated from the data (see
\href{examples}{Example applications} for estimation details). Thus we
arrive at a multivariate negative binomial (MVNB)
\(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).

Table: (\#tab:sims) Identical margin, bivariate simulation
configurations to evaluate correlation matching accuracy and efficiency.

@@

Figure @ref(fig:ch040-bPlot) displays the aggregated bivariate
simulation results. Table @ref(tab:ch040-BiError) contains the mean
absolute error (MAE) in reproducing the desired dependency measures for
the three bivariate scenarios.

\begin{CodeChunk}
\begin{table}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/ch040-bPlot-1}

@@

The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core
Intel Core i9 processor, with all 16 threads employed during
computation. The results for \(d \leq 500\) are fast with all times
(except Kendall) under 3 seconds. Table @ref(tab:ch040-moderateDtab)
displays the total computation time of the algorithms steps 1 through 3,
including estimation step i.

\begin{CodeChunk}

Table: Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.

|Dimension |Correlation type | Total Time (Seconds)|
|:---------|:----------------|--------------------:|
|100       |Pearson          |                0.080|
|100       |Spearman         |                0.036|
|100       |Kendall          |                0.116|
|250       |Pearson          |                0.286|
|250       |Spearman         |                0.371|
|250       |Kendall          |                0.366|
|500       |Pearson          |                2.932|
|500       |Spearman         |                0.176|
|500       |Kendall          |                2.183|

@@

@@

\begin{CodeChunk}
\begin{figure}

@@

@@

\begin{CodeChunk}

@@

\setstretch{1.0}
